---
title: "Lecture 13.1"
subtitle: "Globalization and the China Shock"
author: "Emmanuel Teitelbaum"
footer: "[PSC 2339 Website](https://psc2339.com)"
logo: "images/cpe_logo.png"
format:
  revealjs:
    theme: moon
    transition: fade
    slide-number: true
execute:
  echo: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(fredr)

fredr_set_key(Sys.getenv("FRED_API_KEY"))

# IMPCH: US imports of goods from China, monthly, millions of dollars
# Aggregate to annual and convert to billions
# Exclude the current incomplete year so partial data doesn't distort the chart
china_imports <- fredr(
  series_id = "IMPCH",
  observation_start = as.Date("1990-01-01")
) |>
  mutate(year = year(date)) |>
  filter(year < year(Sys.Date())) |>
  group_by(year) |>
  summarise(imports_bn = sum(value, na.rm = TRUE) / 1000)

```

# America First & The New Protectionism

## "Chief Ripper-Offer"

> "We've been ripped off by every country in the world, but China I would say is the leading candidate for the 'chief-ripper-offer.'"

— Donald Trump, Rose Garden, April 2, 2025

<br>

::: {.fragment}
Trump's "Liberation Day" imposed tariffs of up to **145% on Chinese goods** and double-digit tariffs on virtually every other trading partner. China retaliated with tariffs of 84%+ on US goods. 
:::

## The Retreat from Globalization {.smaller}

- **2018–2019**: First Trump trade war imposed \$360 billion in Section 301 tariffs on Chinese goods (Trade Act of 1974) 
- **2020**: Phase One deal promised no new tariffs if China purchased $200 billion of U.S. goods and services
- **2021–2024**: Biden maintained and expanded Trump-era China tariffs as bipartisan consensus shifted on trade
- **April 2, 2025**: "Liberation Day" imposed sweeping tariffs on nearly every trading partner

::: {.fragment}
**The Political Question** - Why did free trade, long the orthodoxy in both parties, lose its public legitimacy?

The answer may lie in the **China Shock**.
:::

# Globalization: Concepts

## Globalization

<br>

- At the broadest level, globalization is the increasing integration of people, cultures and institutions

- IMF's four aspects: trade; investment; migration; knowledge

- In economics, globalization typically refers to increasing trade and investment

## Benefits of Globalization

<br>

- Increased capital flows
- Sharing of knowledge & technology
- Lower prices for consumers
- Cultural exchange
- New economic opportunities
- Others?

## Challenges of Globalization

<br>

- Exploitation and weak regulation
- High start-up/investment costs
- Localized job loss and displacement
- Immigration pressures
- Erosion of social contracts
- Others?

## Theory of Comparative Advantage {.smaller}

<br>

- **Ricardo (1817)**: a country should specialize in what it can produce at *lower opportunity cost*, even if it is absolutely better at everything

- In a world of US–China trade:
  - China has a comparative advantage in **labor-intensive manufacturing** (abundant, cheap labor)
  - The US has a comparative advantage in **high-skill, technology-intensive sectors**

- Standard prediction: trade liberalization **raises aggregate welfare**
  - US workers displaced from manufacturing *should* reallocate to sectors where the US is competitive
  - Adjustment is assumed to be relatively smooth and fast

- **The critical assumption**: labor and capital move freely across sectors and regions

# The China Shock

## China's Rise {.smaller}

<br>

- In 1989, the *Wall Street Journal* predicted China would be a future *growth laggard*

- Deng Xiaoping's 1992 "Southern Tour" relaunched market reforms:
  - Special Economic Zones (SEZs): 20 in 1991 → 150 by 2010
  - FDI inflows surged from 0.7% of GDP (1980s) to 4.2% (1990s–2000s)

- China's share of world manufacturing exports: **2.3% (1991) → 18.8% (2013)**

- WTO accession in 2001 accelerated the surge

- US imports from China rose by **\$304 billion** between 1991 and 2007

## US Imports from China {.smaller}

```{r}
#| fig-width: 9
#| fig-height: 4.8

ggplot(china_imports, aes(x = year, y = imports_bn)) +
  geom_area(fill = "#e07b54", alpha = 0.2) +
  geom_line(color = "#e07b54", linewidth = 1.4) +
  annotate("segment", x = 2001, xend = 2001, y = 0, yend = 480,
           linetype = "dashed", color = "grey50", linewidth = 0.8) +
  annotate("text", x = 2001.5, y = 490, label = "China joins WTO",
           color = "grey40", size = 3.5, hjust = 0) +
  scale_x_continuous(breaks = seq(1990, 2022, by = 4)) +
  scale_y_continuous(labels = scales::dollar_format(suffix = "B")) +
  labs(x = NULL, y = "US Imports from China (USD billions)",
       caption = "Source: FRED, series IMPCH (US Census Bureau / BEA). Annual totals.") +
  theme_minimal()
```

## The Pre-2000 Consensus — and Its Challenge {.smaller}

**Before China's rise, economists largely agreed**:

1. Trade had *not* been a major cause of US manufacturing employment decline — technology was the main culprit
2. Workers displaced by trade could *readily relocate* to other regions or sectors
3. Any labor-market effects would be *nationally diffuse*, not locally concentrated

. . .

**Autor, Dorn & Hanson (2013/2016) challenge all three**:

- Chinese import competition *did* substantially reduce US manufacturing employment
- The shock was **geographically concentrated** — concentrated in specific communities
- Local adjustment was **slow and painful**, not smooth

## Methodology: Local Labor Markets {.smaller}

<br>

- ADH study **commuting zones (CZs)** — clusters of counties forming local labor markets (722 CZs in the contiguous US)

- **Key insight**: US regions vary in their pre-existing industrial mix → some were far more exposed to Chinese imports than others

- **The identification strategy**: to isolate the *China supply shock* from US demand shifts:
  - Instrument US import growth from China using China's export growth to *other* high-income countries (Australia, Germany, Japan, etc.)
  - If the same Chinese industries were flooding markets in both the US and Germany, it must be a **supply shock** from China — not a reflection of US demand

- This allows a **causal claim**: Chinese import competition → local employment decline

## Key Finding 1: Industry Employment {.smaller}

<br>

- Industries more exposed to Chinese import competition saw substantially larger declines in US manufacturing employment

- **A 1 percentage-point rise in import penetration → 1.3 log-point reduction in industry employment** (Acemoglu et al. 2016)

- The effect accelerated after China's WTO accession (1999–2007):
  - Manufacturing industries contracted by **3.6 log points/year** in this period
  - vs. only 0.3 log points/year during 1991–1999

## Key Finding 2: Regional Losses Don't Wash Out {.smaller}

- Standard trade theory predicts displaced workers *move*, spreading the shock nationally

- ADH find the opposite. In commuting zones more exposed to Chinese imports:

| Outcome | Effect of \$1,000 more per-worker import exposure |
|:---|:---|
| Manufacturing employment share | −0.60 pp |
| Non-manufacturing employment share | −0.18 pp (not sig.) |
| Unemployment rate | +0.22 pp |
| Out of labor force | +0.55 pp |

- Job losses were **not offset** by gains elsewhere in local economies
- Labor mobility was low — especially among less-educated workers

## Key Finding 3: Adjustment Is Slow {.smaller}

<br>

- Standard models assume quick adjustment to trade shocks

- ADH find the adjustment is **remarkably slow**:
  - Wages and labor-force participation remain depressed for **at least a full decade** after the shock begins
  - Exposed workers experience greater job churning and significantly lower lifetime income
  - Government transfer payments (disability, UI) rise substantially in exposed regions

- **Bottom line**: the costs of the China Shock were large, geographically concentrated, and long-lasting — far beyond what standard comparative-advantage models predicted

## Discussion: The Evidence

- What is Autor et al.'s *main* challenge to standard trade theory?

- Why does the methodology matter, e.g. why can't they just compare regions with more vs. fewer manufacturing jobs?

- Does the evidence suggest free trade was a mistake, or that adjustment *support* failed? Is there a difference?

- What does the "no geographic mobility" finding tell us about assumptions built into comparative advantage theory?

# Political Implications

## Did the China Shock Cause the Backlash? {.smaller}

- **Autor, Dorn, Hanson & Majlesi (2020)**: congressional districts more exposed to Chinese imports experienced:
  - Greater **political polarization**
  - Shift toward more **conservative Republican** representatives
  - Erosion of moderate Democrats in manufacturing-exposed districts

- **Colantone & Stanig (2018)**: similar patterns in Europe — regions more exposed to Chinese imports moved toward **nationalist/populist parties**

. . .

This narrative is compelling: trade liberalization hurt specific communities → those communities turned to populist politicians → Trump, Brexit, and the broader wave of economic nationalism

## The Pushback: Mechanism Is Unclear {.smaller}

- The aggregate-to-individual link is **contested**:
  - Research finds weak individual-level evidence that trade-exposed *workers* vote differently than similar workers in less-exposed areas
  - Economic grievance may affect preferences modestly and temporarily (Margalit 2011)
  - **The aggregation problem**: regional patterns may reflect community-level social change, not individual pocketbook voting

- **Alternative explanations** (Inglehart & Norris): "cultural backlash"
  - Identity and values, not just wallets, drive populist voting
  - Affected communities also had high opioid mortality, family dissolution, declining social trust

- **Key question**: Does the China Shock *cause* political change directly through economic grievance, or indirectly through community decline?

## Remaining Questions {.smaller}

<br>

- Is the China Shock → political backlash link *causal*, or were affected regions already culturally/demographically primed for populism?

- Does economic grievance **directly** drive votes, or does it work through:
  - Social network erosion and community decline?
  - Elite cue-taking and media framing (Fox News, etc.)?
  - Identity concerns activated by visible economic loss?

- ADH establish an important **economic** fact — localized, persistent harm from trade — but the **political transmission mechanism** remains an active area of research

# The Bigger Debate

## Two Camps {.smaller}

::: columns
::: {.column width="50%"}
**Free trade's defenders**

- Aggregate gains from trade are real and large
- China's rise lifted 800 million people out of poverty
- US consumers benefited from cheaper goods
- The answer is better adjustment assistance, not protectionism
- Tariffs invite retaliation and raise prices
:::

::: {.column width="50%"}
**China Shock critics**

- Aggregate gains masked severe local costs
- Adjustment assistance was underfunded and ineffective
- Communities — not just workers — were damaged
- Trade policy is inseparable from distributional politics
- The backlash was predictable, and economists ignored the warning signs
:::
:::

## Discussion

- Do you find the "China ripped us off" narrative convincing after reading ADH? What does the evidence actually say?

- If adjustment costs are large and slow, what is the appropriate policy response?
  - Better social insurance? Industrial policy? Trade restrictions?

- Does the empirical evidence *justify* Trump's tariffs? Or is that the wrong question to ask of social science?

- If pocketbook voting doesn't fully explain the Trump coalition, what does?