---
title: "Lecture 13.2"
subtitle: "Globalization"
author: "Emmanuel Teitelbaum"
footer: "[PSC 2339 Website](https://psc2339.com)"
logo: "images/cpe_logo.png"
format:
  revealjs:
    theme: moon
    transition: fade
    slide-number: true
execute:
  echo: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(scales)
library(wid)
library(WDI)
library(wbstats)
library(paletteer)
library(fredr)

fredr_set_key(Sys.getenv("FRED_API_KEY"))

theme_set(theme_minimal())
options(ggplot2.discrete.colour = paletteer::paletteer_d("rcartocolor::Vivid"))
```

# Inequality

## Illustrative statistics from Stiglitz 

<br>

- 20% of wage earners earn as much as bottom 80%
- Top 1% earn in a week what bottom 20% earns in a year
- Top .1% earn in a day what bottom 90% earns in a year
- Walton family controls as much wealth as is owned by bottom 30% of society (~$238 billion)

## Measures We Will Use 

<br>

- Income/wealth shares — what fraction of total income/wealth a group captures
- Gini coefficient — summarizes the whole distribution (0 = perfect equality, 1 = perfect inequality)
- Median household income — measures living standards at the middle; useful for tracking whether the middle is keeping up

## Lorenz Curve and Gini Coefficient

<br>

```{r}
#| label: lorenz
#| fig-height: 3.5
#| fig-width: 6
#| fig-align: center
x <- seq(0, 1, length.out = 500)
df <- data.frame(pop = x, equality = x, lorenz = x^2.5)

ggplot(df, aes(x = pop)) +
  geom_ribbon(aes(ymin = lorenz, ymax = equality), fill = "gray60", alpha = 0.4) +
  geom_ribbon(aes(ymin = 0, ymax = lorenz), fill = "gray40", alpha = 0.3) +
  geom_line(aes(y = equality), color = "gray80", linewidth = 1) +
  geom_line(aes(y = lorenz),
            color = paletteer::paletteer_d("rcartocolor::Vivid")[1], linewidth = 1) +
  geom_segment(x = 0.4, xend = 0.4, y = 0, yend = 0.4^2.5,
               linetype = "dashed", color = "gray60") +
  geom_segment(x = 0, xend = 0.4, y = 0.4^2.5, yend = 0.4^2.5,
               linetype = "dashed", color = "gray60") +
  geom_segment(x = 0.8, xend = 0.8, y = 0, yend = 0.8^2.5,
               linetype = "dashed", color = "gray60") +
  geom_segment(x = 0, xend = 0.8, y = 0.8^2.5, yend = 0.8^2.5,
               linetype = "dashed", color = "gray60") +
  annotate("text", x = 0.50, y = 0.44, label = "A", size = 7,
           color = "white", fontface = "bold") +
  annotate("text", x = 0.75, y = 0.14, label = "B", size = 7,
           color = "white", fontface = "bold") +
  scale_x_continuous(labels = percent, breaks = c(0, 0.2, 0.4, 0.8, 1.0)) +
  scale_y_continuous(labels = percent,
                     breaks = c(0, 0.10, 0.40, round(0.8^2.5, 2), 1.0)) +
  labs(x = "Population (poorest to richest)", y = "Cumulative income share")
```

:::{style="font-size: 0.6em"}
The diagonal is the **line of equality** — every X% of the population earns Y% of income. The bowed curve is the **Lorenz curve** (the actual distribution). **Gini = A/(A+B)**: ranges from 0 (perfect equality) to 1 (perfect inequality).
:::

::: {.notes}
The diagonal line represents perfect equality — every x% of the population earns x% of income. The Lorenz curve shows the actual distribution, bowed below the line of equality. Region A is the gap between the two; region B is the area under the Lorenz curve. The Gini coefficient is A/(A+B): the larger A is relative to the total triangle, the more unequal the distribution. It ranges from 0 (perfect equality) to 1 (perfect inequality).
:::


## Income vs. Wealth Inequality

<br>

- Income: what people earn from work and *returns* from investments like stocks, bonds and investment properties 
- Wealth: value of everything a person or family owns minus any debts
	- Net worth: marketable assets minus debts
	- Financial wealth: non-home wealth
	
# Wealth Inequality

## Bottom 90% vs. Top 10% Wealth Shares (U.S.)

```{r}
#| label: us-ineq3
download_wid(
	indicators = "shweal", # Shares of personal wealth
	areas = "US", # In the U.S.
	perc = c("p0p90", "p90p100"), # Bottom 90% and top 10%
	ages = "992", 
	pop = "j"
  ) %>% 
  ggplot(aes(x=year, y=value, color=percentile)) +
  	geom_line() +
  	scale_y_continuous(label=percent) +
  	scale_color_discrete(labels=c("p0p90"= "bottom 90%", "p90p100"="top 10%")) +
  	labs(title = "Bottom 90% and Top 10% Personal Wealth Shares in the U.S.", 
  		 caption = "Source: World Inequality Database",  
  		 x = "Year", y = "Top Share")
```

## The Problem of the 1% 

```{r}
#| label: us-ineq4
download_wid(
	indicators = "shweal", # Shares of personal wealth
	areas = "US", # In the U.S.
	perc = c("p0p50", "p99p100"), # Bottom 50% and top 1%
	ages = "992", 
	pop = "j"
) %>% 
  ggplot(aes(x=year, y=value, color=percentile)) +
  	geom_line() +
  	scale_y_continuous(label=percent) +
  	scale_color_discrete(labels=c("p0p50"="bottom 50%", "p99p100"="top 1%")) +
  	labs(title = "Bottom 50% and Top 1% Personal Wealth Shares in the U.S.", 
  		 caption = "Source: World Inequality Database",  
  		 x = "Year", y = "Top Share")
```

## Decomposing the 1%

```{r}
#| label:  us-ineq5
download_wid(
	indicators = "shweal", # Shares of personal wealth
	areas = "US", # In France
	perc = c("p99p100", "p99.5p100", "p99.9p100", "p99.99p100"), # Shares of 1%
	ages = 992, 
	pop = "j"
) %>%
  ggplot(aes(x=year, y=value, color=percentile)) +
  	geom_line() +
  	scale_y_continuous(label=percent) +
  	scale_color_discrete(
  	  breaks = c("p99p100", "p99.5p100", "p99.9p100", "p99.99p100"), # to change order
  	  labels=c("p99p100"="top 1%", "p99.5p100"="top 0.5%",
  	           "p99.9p100"="top 0.1%", "p99.99p100"="top 0.01%")
  	           ) + 
      labs(title = "Top Wealth Shares in the U.S.", 
  		  caption = "Source: World Inequality Database",  
  		  x = "Year", y = "Top Share")
```

::: {.notes}
Top 1% as a whole see increased wealth shares of about 15% from 1975 to now, 
while top .5% doubles its wealth share; .1% and .01% tripple their wealth shares. 
:::

## Wealth Shares in France

```{r}
#| label: france-ineq
download_wid(
	indicators = "shweal", # Shares of personal wealth
	areas = "FR", # In France
	perc = c("p90p100", "p99p100") # Top 10% and top 1%
) %>%
  ggplot(aes(x=year, y=value, color=percentile)) +
  	geom_line() +
  	scale_y_continuous(label=percent) +
  	scale_color_discrete(labels=c("p90p100"="top 10%", "p99p100"="top 1%")) +
  	labs(title = "Top 1% and Top 10% Personal Wealth Shares in France",
  		 caption = "Source: World Inequality Database",
  		 x = "Year", y = "Top Share")
```

## Piketty: Why Wealth Concentrates

<br>

- Thomas Piketty's central argument: **r > g**
  - *r* = rate of return on capital (dividends, interest, rent, capital gains)
  - *g* = rate of economic growth
- When r > g, those who own wealth accumulate faster than the economy grows
- Historical averages: r ≈ 4–5%, g ≈ 1–2% over the long run

## The Mechanism

<br>

- Owners of capital reinvest returns → wealth compounds over time
- Workers depend on wage growth, which roughly tracks *g*
- Over time, capital's share of total income rises relative to labor's share
- Inherited wealth comes to dominate over earned income
  - Piketty: "the past devours the future"

## Exercise: Testing Piketty {.smaller}

- Go to [wid.world](https://wid.world) → Country graphs → Wealth inequality
- Select a country: Canada, United Kingdom, Germany, France, Sweden, China, or India
- Look at top wealth shares over time and discuss:
  - Has wealth concentrated at the top in your country?
  - Does the trend fit Piketty's prediction — slow, steady concentration driven by capital returns?
  - Or does the pattern look different? What might explain that?

## Limitations of r > g

<br>

- Predicts slow, gradual concentration — but U.S. inequality rose sharply after ~1980
- Most Americans have little non-home wealth to compound, so r > g doesn't help them
- Better at explaining *wealth* concentration than *income* inequality
  - We need a different theory for that

# Income Inequality

## United States Gini Coefficient

```{r}
#| label: us_gini
WDI(indicator="SI.POV.GINI", country='USA', start=1990) %>%
  ggplot(aes(year, SI.POV.GINI)) + 
    geom_line(color = paletteer::paletteer_d("rcartocolor::Vivid")[1]) +
  	labs(title = "Gini Index of Inequality in the U.S.", 
		 caption = "Source: World Bank",  
		 x = "Year", y = "Gini Index")
```

## United States Median Household Income

```{r}
#| label: median_income
fredr(series_id = "MEHOINUSA672N", observation_start = as.Date("1984-01-01")) %>%
  ggplot(aes(date, value)) +
  geom_line(color = paletteer::paletteer_d("rcartocolor::Vivid")[1]) +
  scale_y_continuous(label = comma) +
  labs(title = "Real Median Household Income in the United States",
       caption = "Source: U.S. Census Bureau via FRED",
       x = "Year", y = "2021 CPI-U-RS Adjusted Dollars")
```

## Bottom 90% vs. Top 10% Income Shares (U.S.)

```{r us-ineq1, out.width="100%", echo = FALSE}
download_wid(
	indicators = "sptinc", # Shares of pre-tax national income
	areas = "US", # In the U.S.
	perc = c("p0p90", "p90p100"), # Bottom 90% and top 10%  
	pop = "j"
) %>%
  ggplot(aes(x=year, y=value, color=percentile)) +
  	geom_line() +
  	scale_y_continuous(label=percent) +
  	scale_color_discrete(labels=c("p0p90"="bottom 90%", "p90p100"="top 10%")) +
  	labs(title = "Bottom 90% and Top 10% Income Shares in the U.S.", 
  		 caption = "Source: World Inequality Database",  
  		 x = "Year", y = "Top Share")
```

## United States Average Incomes

```{r}
#| label: us-ineq2
download_wid(
	indicators = "aptinc", # Average pre-tax national income
	areas = "US", # In the U.S.
	perc = c("p0p50", "p50p60", "p60p70", "p70p80", "p80p90", "p90p100"), # Percentiles
	pop = "j",
	ages = "992",
	years = 1970:2022
) %>%
  ggplot(aes(x=year, y=value, color=percentile)) +
	  geom_line() +
  	scale_y_continuous(label=comma) +
  	scale_color_discrete(
  	  labels=c("p0p50"="bottom 50%", "p50p60"="6th decile", "p60p70"="7th decile", "p70p80"="8th decile", "p80p90"="9th decile", "p90p100"="top 10%")) +
	  labs(title = "Average Income by Percentile", 
	    caption = "Source: World Inequality Database", x = "Year", y = "Dollars")
```

## Stiglitz: Rent-Seeking and Market Power

- Joseph Stiglitz argues inequality stems from **rent-seeking**
- Economic rent = income earned by *capturing* value rather than *creating* it
  - Monopoly pricing, financial speculation, control over intellectual property
- Key insight: market rules are not natural — they are shaped by those with power
  - Who benefits from deregulation? From weak antitrust enforcement?

## Market Power and Worker Wages

- **Monopsony**: when employers dominate local labor markets, they can suppress wages
  - Workers paid less than the value they produce
- Non-compete clauses trap workers and reduce their bargaining power
- Fissured workplace: outsourcing, franchising, and gig economy shift risk onto workers
- Concentrated industries (healthcare, tech, retail) → lower wages for workers

## Executive Pay vs. Worker Pay

- CEO pay has grown ~940% since 1978; typical worker pay ~12%
- Executives sit on each other's boards → mutual back-scratching on compensation
- Shareholder primacy shifts gains to executives and shareholders at workers' expense
- Stock buybacks and options concentrate rewards at the very top
- This is Stiglitz's core target: inequality *within* firms, not just between them

## Stiglitz's Remedies {.smaller}

<br>

- **Revive antitrust enforcement** — break up existing monopolies, block anticompetitive mergers
- **Broaden the standard** — move beyond consumer prices to consider worker welfare, innovation, and market structure
- **Regulate platforms** — prevent companies from competing on their own marketplace (Amazon selling against third-party sellers)
- **Reform corporate governance** — give workers a seat at the table; rein in executive pay
- **Reduce political influence of corporations** — money in politics perpetuates the rules that allow rent-seeking

## Debate: Fixing Market Power 

- Pick one of the major tech platforms: Meta, Amazon, Apple, Google, or Microsoft
- Stiglitz proposes several remedies — which would you apply to your company, and why?
  - Break it up?
  - Regulate it as a platform?
  - Reform its governance?
  - Something else?
- Who wins and who loses from your proposed remedy?

## Lina Khan and the New Antitrust {.smaller}

<br>

> "Anti-monopoly is a governing philosophy that broadly views concentrations of power as a threat to freedom, recognizing that tyranny comes in many guises. Just as the Constitution creates checks and balances in our government, safeguarding against concentrated political control, anti-monopoly laws create checks against concentrations of economic power."

— Lina Khan, former FTC Chair (2021–2025)

<br>

::: {.fragment}
- Khan argued the **consumer welfare standard**, antitrust's focus on keeping prices low, is blind to how platforms accumulate and abuse market power
- How does Khan's argument follow from Stiglitz? Should we think of concentrated economic power as a threat to freedom?  
:::

# Conclusion

## The Bigger Picture {.smaller}

<br>

- Piketty and Stiglitz give us two complementary mechanisms:
  - **Wealth inequality**: capital compounds faster than economies grow → the wealthy pull away
  - **Income inequality**: market power suppresses wages and inflates executive pay → workers fall behind
- Neither theory alone is sufficient because inequality is overdetermined
  - Multiple forces reinforced each other from the 1980s onward
- Inequality is not an accident of markets
- It reflects the rules of the game, and rules can be changed

## Other Forces Worth Noting {.smaller}

<br>

These factors amplified the Piketty/Stiglitz dynamics rather than replacing them:

- **Economic globalization** — trade competition suppressed wages in exposed sectors; financialization expanded returns to capital (Piketty)
- **The Reagan revolution** — deregulation, removal of capital controls, and cuts to top marginal tax rates weakened redistribution and widened the gap between r and g (both)
- **Technology and skill change** — reduced demand for low-skilled workers, increased returns to education (Stiglitz)
- **Decline of unions and left parties** — eroded worker bargaining power, accelerating wage stagnation (Stiglitz)
- **Weakening redistribution** — lower minimum wages and a thinner safety net allowed market inequality to translate more directly into living standards

