---
title: "Lecture 12.2"
subtitle: "Central Bank Independence"
author: "Emmanuel Teitelbaum"
footer: "[PSC 2339 Website](https://psc2339.com)"
logo: "images/cpe_logo.png"
format:
  revealjs:
    theme: moon
    transition: fade
    slide-number: true
execute:
  echo: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(ggrepel)

# Data from Alesina & Summers (1993), 16 OECD countries, 1955-1988
# CBI index from Table 1 (average of BP and GMT scales)
# Inflation and GNP growth approximated from Figures 1a and 2a

cbi_data <- tibble(
  country = c("AUS", "BEL", "CAN", "DEN", "FRA", "GER", "ITA", "JAP",
              "NET", "NOR", "NZ",  "SPA", "SWE", "SWI", "UK",  "USA"),
  cbi     = c(2.0,   2.0,   2.5,   2.5,   2.0,   4.0,   1.75,  2.5,
              2.5,   2.0,   1.0,   1.5,   2.0,   4.0,   2.0,   3.5),
  avg_inf = c(6.5,   4.1,   4.6,   6.8,   6.6,   3.0,   7.5,   5.1,
              4.3,   6.7,   7.9,   9.3,   6.5,   3.2,   6.7,   4.2),
  avg_gnp = c(3.8,   3.1,   4.0,   3.1,   3.6,   3.3,   4.0,   6.5,
              3.3,   3.6,   1.5,   4.2,   2.5,   2.3,   2.2,   3.2)
)

dark_theme <- theme_minimal(base_size = 13) +
  theme(
    plot.background  = element_rect(fill = "#191919", color = NA),
    panel.background = element_rect(fill = "#191919", color = NA),
    panel.grid.major = element_line(color = "#333333"),
    panel.grid.minor = element_line(color = "#2a2a2a"),
    text             = element_text(color = "white"),
    axis.text        = element_text(color = "white"),
    axis.title       = element_text(color = "white"),
    plot.caption     = element_text(color = "grey60", size = 9)
  )
```

# The Fed Under Fire

## "Mr. Too Late"

> "There can almost be no inflation, but there can be a SLOWING of the economy unless **Mr. Too Late**, a major loser, lowers interest rates, **NOW**."

— Donald Trump, Truth Social, April 21, 2025

<br>

::: {.fragment}
Trump subsequently threatened to fire Fed Chair Jerome Powell. Markets fell ~2.4% the day of the post; the dollar also fell. Powell said he would not resign and Trump later backed off: "I have no intention of firing him."
:::

## What Does the Fed Do?

<br>

- Sets the **federal funds rate** (the benchmark interest rate)
- Controls the money supply and credit conditions
- Acts as lender of last resort to the banking system
- Mandated by Congress to pursue **price stability** *and* **maximum employment**
- Operates independently of the White House (by design)

## An Old Debate

- Politicians have *always* been tempted to pressure central banks
  - Lower rates → cheaper credit → short-term boost to growth and jobs
  - But expansionary monetary policy can cause inflation
- **Political business cycle** theory (Nordhaus 1975): governments inflate before elections
- Nixon's pressure on Fed Chair Arthur Burns before 1972 election (a classic case)
- Should monetary policy be insulated from politics?

# Alesina & Summers Framework

## Two Central Questions

::: {.fragment}
1. Does **central bank independence (CBI)** reduce inflation?
2. If so, is there a **cost** to real economic performance, e.g. growth & unemployment?
:::

::: {.incremental}
- If the answer to (1) is **yes** and (2) is **no** — that is a **"free lunch"**
- Alesina and Summers argue this is exactly what the data show
:::

## Why Might Politicians Want Easy Money?

- Real output *can* be temporarily boosted by loose monetary policy
  - Lower rates → cheaper borrowing → more investment and consumption
- **Political business cycles**: rate cuts before elections inflate short-term growth
- Classic example: Nixon pressuring Fed Chair Aurthor Burns before 1972 presidential election

## The Problem

<br>

- **Dynamic inconsistency** (Kydland & Prescott 1977; Barro & Gordon 1983)
  - Governments face a short-term temptation to inflate: boost output and cut unemployment before elections
  - But if people *expect* inflation, they demand higher wages → prices rise with no lasting output gain
  - A government that cannot commit to low inflation ends up with high inflation *and* no output benefit

## The Solution: Central Bank Independence

<br>

  - Delegate monetary policy to a conservative, independent central banker
  - Independent central bankers are more inflation-averse than the median voter
  - Can *credibly commit* to price stability in a way that elected governments cannot
  - Political insulation = credible commitment = lower inflation expectations 

## The Counter-Argument

<br>

  - Politically responsive central banks may better stabilize output
  - Independent central banks may tolerate higher unemployment to fight inflation 
  - Primarily research by Rogoff (1985) 

## Measuring Central Bank Independence {.smaller}

<br>

- Not straightforward because independence is multi-dimensional
- Two main indices:
  - **Bade-Parkin (BP)**: 1–4 scale based on *political* independence — relationship with the executive, appointment procedures, government representation on the board
  - **Grilli-Masciandaro-Tabellini (GMT)**: adds *economic* independence — can the government force the central bank to finance deficits?
- Alesina & Summers average the two into a single 1–4 scale (higher = more independent)
- Sample: **16 OECD countries, 1955–1988**

## Who Is Most (and Least) Independent? {.smaller}

<br>

| Least Independent | CBI | Most Independent | CBI |
|:---|:---:|:---|:---:|
| New Zealand | 1.0 | Germany | 4.0 |
| Spain | 1.5 | Switzerland | 4.0 |
| Italy | 1.75 | United States | 3.5 |
| Australia, Belgium, France, Norway, Sweden, UK | 2.0 | Japan, Canada, Netherlands, Denmark | 2.5 |

<br>

## Discussion: Before We Look at the Data

<br>

- Given these rankings, what do you *expect* the relationship to be between CBI and inflation?

- What about CBI and economic growth?

- Does the ranking of any country surprise you?

# The Evidence

## CBI and Average Inflation {.smaller}

```{r}
#| fig-width: 9
#| fig-height: 5.2

ggplot(cbi_data, aes(x = cbi, y = avg_inf, label = country)) +
  geom_smooth(method = "lm", se = FALSE, color = "#e07b54", linewidth = 0.9) +
  geom_point(color = "coral4", size = 3) +
  geom_text_repel(color = "darkgreen", size = 3.5,
                  box.padding = 0.5, point.padding = 0.3,
                  segment.color = "grey50", max.overlaps = 20) +
  scale_x_continuous(limits = c(0.5, 4.7), breaks = 1:4) +
  scale_y_continuous(limits = c(2, 11)) +
  labs(x = "Index of Central Bank Independence",
       y = "Average Inflation (%)",
       caption = "Source: Alesina & Summers (1993). Data: 1955–1988.") +
  theme_minimal()
```

## CBI and Average Real GNP Growth {.smaller}

```{r}
#| fig-width: 9
#| fig-height: 5.2

ggplot(cbi_data, aes(x = cbi, y = avg_gnp, label = country)) +
  geom_smooth(method = "lm", se = FALSE, color = "#e07b54", linewidth = 0.9) +
  geom_point(color = "coral4", size = 3) +
  geom_text_repel(color = "darkgreen", size = 3.5,
                  box.padding = 0.5, point.padding = 0.3,
                  segment.color = "grey50", max.overlaps = 20) +
  scale_x_continuous(limits = c(0.5, 4.7), breaks = 1:4) +
  scale_y_continuous(limits = c(0.5, 8)) +
  labs(x = "Index of Central Bank Independence",
       y = "Average Real GNP Growth (%)",
       caption = "Source: Alesina & Summers (1993). Data: 1955–1988.") +
  theme_minimal()
```

## What Do You See?

<br>

- What pattern do you see in the **inflation chart**?

- What pattern do you see in the **GNP growth chart**?

- What is the implication if *both* patterns are true simultaneously?

- Any outliers that stand out? What might explain Japan? New Zealand?

## The "Free Lunch" Result

- **CBI strongly associated with lower inflation** — near-perfect negative correlation
  - Also holds for inflation *variability* (more independent banks → more stable prices too)

- **CBI has no measurable effect on real output** — flat relationship with GNP growth, unemployment, and real interest rates

- **Implication**: CBI delivers price stability at *no cost* to economic performance (free lunch!)

# Subsequent Research

## What's Wrong with the Alesina & Summers Study? {.smaller}

- **Small N, no controls**: only 16 countries; simple scatter plots with no regression or confounders
  - What if something *else* — trade openness, fiscal discipline — explains both higher CBI and lower inflation?

- **Endogeneity**: does CBI *cause* low inflation, or do countries with pre-existing low-inflation preferences simply grant more independence?
  - Germany: 1920s hyperinflation trauma → public aversion to inflation → independent Bundesbank
  - The correlation could run in reverse

- **De jure ≠ de facto**: indices measure central bank *laws*, not actual behavior (laws can be ignored)

- **Historical period**: all data from 1955–1988; the world has changed substantially since

## What Has Research Found Since? {.smaller}

<br>

- **Klomp & de Haan (2010)** — meta-regression of 59 studies: the negative CBI–inflation link is real even correcting for publication bias, but the size of the effect varies across specifications

- **Bodea & Hicks (2015)** — panel data with instrumental variables: CBI lowers inflation and the "free lunch" holds, *but mainly in democracies* where legal commitments are credible

- **Acemoglu et al. (2008)** — most important challenge: CBI is largely **endogenous to political institutions**; where executive power is unconstrained, CBI reform is ineffective because politicians find other ways to inflate (fiscal dominance, financial repression)

- **Bottom line**: for developed democracies, Alesina & Summers holds reasonably well with modern methods; for developing countries, institutional context is everything

## Developing Countries: The De Jure / De Facto Gap {.smaller}

<br>

- In many developing countries, central banks are legally independent but not *actually* independent

- **Cukierman et al. (1992)**: use central bank **governor turnover** as a proxy for de facto independence
  - High turnover = governors being fired = low real independence
  - Legal CBI predicts little in developing countries; actual behavior predicts more

- **Garriga (2016)**: 182-country panel shows that CBI reduces inflation in developing countries, but only where **rule of law** is strong

- **Key insight** (Acemoglu et al.): CBI may be a *symptom* of good institutions rather than a cause of good outcomes

## Post-2008: New Pressures {.smaller}

<br>

- **Quantitative easing (QE)**: central banks purchased government bonds at massive scale to fight the financial crisis and COVID recession
  - Blurs the monetary/fiscal line
  - Risk of **fiscal dominance**: central banks that hold large amounts of government debt may be unable to raise rates without triggering sovereign debt crises

- **Distributional politics**: near-zero rates and asset purchases benefited asset-holders and debtors; savers and pensioners lost
  - Made central banks politically visible in a new way, i.e. harder to claim "apolitical" status

- **Post-COVID inflation surge**: central banks were slow to raise rates in 2021–22; accusations of political influence resurfaced

# A Natural Experiment: Turkey

## When Central Bank Independence Breaks Down {.smaller}

- President Erdogan dismissed **six central bank governors** between 2016 and 2023
- Governors were fired for resisting pressure to cut rates despite rising inflation
- Erdogan's unorthodox theory: high interest rates *cause* inflation (mainstream economics says the opposite)

. . .

| Year | Inflation |
|:-----|----------:|
| 2019 | ~11% |
| 2021 | ~21% |
| 2022 | **~85%** |
| 2023 | ~65% (after policy reversal) |

. . .

- Turkey eventually reversed course; a new governor hiked rates and inflation fell
- But credibility damage was lasting (near perfect test of A&S logic)

# The Bigger Debate

## Two Camps {.smaller}

::: columns
::: {.column width="50%"}
**CBI as technocratic solution**

- Elected politicians have short-term incentives; monetary policy requires a long view
- Evidence: CBI lowers inflation at no output cost (for developed economies)
- Turkey shows what happens when independence is removed
- Insulating the central bank from electoral cycles is simply good institutional design
:::

::: {.column width="50%"}
**CBI as democratic problem**

- Unelected officials make decisions with large distributional consequences
  - High rates hurt debtors, workers, homebuyers
  - Low rates hurt savers, pensioners
- "Insulated from politics" often means insulated from accountability
- QE and the 2008 crisis showed central banks are never truly apolitical
:::
:::

## Discussion

<br>

- Should central banks be independent of political control?

- Who wins and who loses when interest rates are *high*? When they are *low*?

- Does the Trump-Powell conflict change your view? Does the Turkey case?

- Is there a middle ground between full independence and full political control?
