---
title: "Lecture 7.1"
subtitle: "Washington Consensus Reforms"
author: "Emmanuel Teitelbaum"
footer: "[PSC 2339 Website](https://psc2339.com)"
logo: "images/cpe_logo.png"
format:
  revealjs:
    theme: moon
    transition: fade
    slide-number: true
execute:
  echo: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(WDI)
library(RColorBrewer)
```

# The Washington Consensus

## What Is It?

-   A policy framework associated with market-oriented reform in the
    1980s-1990s
-   Originally a list of policy prescriptions (often linked to IMF /
    World Bank reform programs)
-   Not the same thing as "all neoliberalism" in every context
-   Important in this course because it shaped how crises and reform
    were understood in many countries

## 10 Policy Prescriptions {.smaller}

1.  Fiscal discipline
2.  Public expenditure priorities (investment vs consumption)
3.  Tax reform (broader base, lower marginal rates)
4.  Financial liberalization
5.  Competitive exchange rates
6.  Trade liberalization
7.  Foreign direct investment (remove barriers)
8.  Privatization of state-owned enterprises
9.  Deregulation of markets
10. Secure property rights

## Reform Elements 

-   Stabilization
    -   Fiscal discipline, inflation control, exchange rate policy
-   Liberalization / opening markets
    -   Trade, finance, FDI, deregulation
-   State restructuring
    -   Privatization, public spending reprioritization, tax reform
-   Longer-run institutional reform
    -   Property rights, regulatory capacity, state capability

# Crisis and Reform

## Why Crisis Creates Reform Windows

-   Recession, inflation, shortages, and debt make the status quo
    politically costly
-   Balance of payments crises create immediate external financing
    pressure
-   International lenders often condition support on reform packages
-   Reformers frame crisis as proof that "the old model" has failed

## Elements of an Economic Crisis

<br>

-   Recession / unemployment
-   Inflation (sometimes hyperinflation)
-   Balance of payments crisis
-   Debt crisis
-   Financial crisis / banking instability

## 

::: columns
::: {.column width="50%"}
### Eastern Europe

- Collapse of communist systems
- Price distortions, shortages, inflation
- Push for rapid transition to market economies
- Big external role for international experts / IFIs
:::

::: {.column width="50%"}
### India (1991)

- Balance of payments crisis
- Trade shock after Soviet collapse
- Oil shock from Gulf War
- Capital flight and currency pressure
- Reform under crisis, but without full system collapse
:::
:::

# Reform Packages

## Reform #1: Stabilization

-   Goal: stop inflation / currency crisis from spiraling
-   Typical tools
    -   Cut or restrain public spending
    -   Raise revenue / taxes
    -   Devalue currency
    -   Tighten monetary policy
-   Political problem: immediate pain, delayed gains

## Reform #2: Liberalization

<br>

-   End controls (prices, licenses, quotas, reservations)
-   Reduce tariffs and trade barriers
-   Expand FDI access
-   Financial sector reforms
-   Logic: competition, investment, innovation, and export growth

## Reform #3 Privatization 

<br>

-   Transfer state assets to private actors
-   Improve incentives and efficiency (in theory)
-   Raise state revenue

## Reform #4 Structural adjustment

<br>

-   Longer-term restructuring of how the economy is governed
-   Labor, land, capital markets; industrial and investment policy
-   Core debate: market reform without state capacity can backfire

# Shock Therapy vs. Gradualism

## 

::: columns
::: {.column width="50%"}
### Shock Therapy

-   Rapid, broad reforms
-   Front-load stabilization and liberalization
-   Goal: break old system quickly
-   Commonly associated with Eastern Europe
:::

::: {.column width="50%"}
### Gradualism

-   Sequenced reforms over time
-   Build institutions while opening markets
-   Slower on politically costly reforms
-   India is a key example
:::
:::

## India vs. Eastern Europe (Sequencing)

-   India moved relatively quickly on trade / investment liberalization
-   India moved more slowly on privatization and labor reform
-   Eastern Europe often attempted broader reforms much faster
-   Key question for us:
    -   When does speed solve problems?
    -   When does speed outpace institutions?

## Risk: "Partial Reform Equilibrium"

<br>

-   Early winners from reform may block later reforms
-   Privatization without strong institutions can create asset capture
-   Result can be a hybrid system:
    -   Neither effective markets nor accountable states
-   This is one reason "reform" outcomes vary so much

# Outcomes and Debate

## Growth Rates in India and China

```{r growth_graph, echo = FALSE}
flfp <- WDI(indicator="NY.GDP.MKTP.KD.ZG", country=c('IN','CN'), start=1990) 
    
ggplot(flfp, aes(year, NY.GDP.MKTP.KD.ZG, color=country)) + geom_line(linewidth=1) +
	xlab('Year') + ylab('GDP Growth') + scale_color_brewer(palette="Dark2") +
  labs(
    caption = "World Bank national accounts data, and OECD National
Accounts data files"
  )
```

## What Growth Rates Tell Us

<br>

-   Shows growth performance over time (and volatility)
-   Does not tell us by itself:
    -   Which reforms caused growth
    -   Who benefited
    -   What social costs occurred during transition
-   This is why we need comparative and historical analysis

# Legacies of Reform 

## Reading Focus (Archibong et al.)

<br>

-   Reading for today:
    [Washington Consensus Reforms and Lessons for Economic Performance in
    sub-Saharan Africa](https://www.brookings.edu/wp-content/uploads/2021/02/Washington-Consensus-Reforms.pdf)
-   Main question:
    -   What were the long-run effects and limits of Washington
        Consensus-style reforms in Africa?
-   Working distinction:
    -   Reform adoption is not the same as reform success

## Discussion Questions (Small Groups)

-   What is the authors' main argument about the legacy of reforms in
    Africa?
-   Which reforms seem most important in their account, and why?
-   What role do state capacity and political institutions play?
-   How do the authors treat short-run pain vs. long-run gains?
-   What are the limits of the evidence (quantitative and case-based)?

## Group Report-Out

<br>

-   Each group: one claim from the reading the class should take
    seriously
-   Each group: one claim you are skeptical about (and why)
-   Connect back to today:
    -   Does your evidence support shock therapy, gradualism, or "it
        depends"?

# Contemporary Application

## Lebanon: Crisis and Reform in Practice

-   [When robbing a bank seems perfectly justified](https://www.nytimes.com/2023/02/28/opinion/lebanon-crisis-economy-bank-robbers.html)
-   Ongoing recovery / reform context:
    [World Bank on Lebanon](https://www.worldbank.org/en/country/lebanon/overview#1)
-   Use Lebanon as an application exercise:
    -   Crisis diagnosis
    -   Reform package design
    -   Political feasibility

## Lebanon Discussion {.smaller}

<br>

-   Is Lebanon's crisis mainly a macroeconomic crisis, a banking /
    financial crisis, a governance crisis, or all three?
-   Which Washington Consensus-style reforms might help in Lebanon?
-   Which reforms could deepen social pain without protections?
-   What sequencing would you propose:
    -   Stabilization first?
    -   Institutional repair first?
    -   Both in parallel?


