---
title: "Lecture 2.2"
subtitle: "Modernization Theory"
author: "Emmanuel Teitelbaum"
footer: "[PSC 2339 Website](https://psc2339.com)"
logo: "images/cpe_logo.png"
format:
  revealjs:
    theme: moon
    transition: fade
    slide-number: true
    css: custom.css
execute:
  echo: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(vdemdata)
library(plotly)
library(WDI)

theme_set(theme_minimal())

source("helper/lecture-2.2.R")
```

# Measuring Wealth

## GDP and GDP per Capita

-  Gross Domestic Product (GDP)
    -   Total value of goods and services produced within a country
-   GDP per capita
    -   GDP divided by total population
- Generally speaking, we use GDP per capita to measure wealth of
  individual countries 
- Whereas GDP may be more of a measure of power or influence 

## World's Largest Economies 

<br>

```{r}
#| label: wealthiest1

ggplot(biggest_ppp, aes(y = fct_reorder(country, gdp_ppp), x = gdp_ppp)) +
  geom_col(fill = "red3") +
  scale_x_continuous(labels = function(x) paste0("$", x/1000000000000, " ", "Trillion")) +
  labs(x = "\nGDP, PPP (constant 2017 International $)", y = NULL,
       title = "The fifteen largest economies by GDP in terms of PPP",
       subtitle = "Most recent available year",
       caption = "\nSource: The World Bank, World Development Indicators")
```

## World's Wealthiest Countries

<br>

```{r}
#| label: wealthiest2

ggplot(wealthiest_pc_ppp, aes(y = fct_reorder(country, gdp_pc_ppp), x = gdp_pc_ppp)) +
  geom_col(fill = "red3") +
  scale_x_continuous(labels = scales::label_dollar()) +
  labs(x = "\nGDP per capita, PPP (constant 2017 International $)", y = NULL,
       title = "The fifteen wealthiest countries by GDP per capita (PPP)",
       subtitle = "Most recent available year",
       caption = "\nSource: The World Bank, World Development Indicators") 
```

# Measuring Democracy

## The V-Dem Institute

-   Varieties of Democracy (V-Dem)
-   Polyarchy score
    -   Robert Dahl
    -   Polyarch--form of government in which power is vested in
        multiple people
-   V-Dem's polyarch index is based on concept of electoral democracy

## Polyarchy (Electoral Democracy)

<br>

**Polyarchy:** "seeks to embody the core value of making rulers responsive
to citizens, achieved through electoral competition...\[P\]olitical and
civil society organizations can operate freely, elections are clean..and
affect the composition of the chief executive."

## Other V-Dem Indices

- **Liberal:** checks on executive power, rights protections, rule of law
- **Participatory:** citizen participation beyond elections
- **Deliberative:** reasoned public debate and justification
- **Egalitarian:** equal political access and influence

## Most Democratic Countries (Polyarchy)

<br>

```{r}
#| label: polyarchy_top15
ggplot(polyarchy_top15, aes(y = fct_reorder(country, polyarchy), x = polyarchy)) +
  geom_col(fill = "steelblue1") +
  scale_x_continuous(labels = scales::label_number(accuracy = 0.01)) +
  labs(x = "\nPolyarchy score", y = NULL,
       title = paste0("Top 15 countries by polyarchy (", latest_vdem_year, ")"),
       caption = "\nSource: V-Dem")
```

# Modernization Theory

## What is Modernization Theory?

-   Based on observation by Lipset (1959)
-   Noted that democracy is highly correlated with development
-   Took on guise of causal argument
    -   Wealth → democaracy

## Some Stylized Facts

-   Wealthier countries tend to be more democratic
-   Relationship tends to be stronger before WWII
-   There also tend to be big regional differences
    -   In some regions the relationship is evident
    -   In others, not as much

## Modernization Before 1945

<br>

<div class="center-plotly">
```{r}
#| label: modernization1

modernization_plot1
```
</div>

## Modernization in the Post-War Period

<br> 

<div class="center-plotly">
```{r}
#| label: modernization2

modernization_plot2
```
</div>

## "Endogenous Democratization"

-   Democracy is a direct *consequence* of economic development
-   Przeworski's term for modernization theory
    -   *Endogenous*--coming from within
    -   Democracy is a result of wealth
    
## Proposed Mechanisms

-   Industrialization
-   Urbanization
-   Rise of middle class
-   Education

## Exogenous Democratization

-   *Exogenous*--resulting from an external cause
-   Wealth does not cause democracy
-   But democracy is more easily sustained in wealthy countries
-   Przeworski's argument against modernization theory

## Democratic Survival and Income {.smaller}

<br>

> “No democracy has ever fallen in a country with a per capita income higher than that of Argentina in 1975.”

— Adam Przeworski et al., *Democracy and Development* (2000)

<br>

Empirical threshold
- ≈ $6,000 GDP per capita (1990 PPP dollars)
- Above this level, democracies are observed to be ‘immortal’ in the data

<br>

**Key point:** income predicts **democratic survival**, not democratic transition

# Breaking it Down

## Democracy and Development by Region

<br>

<div class="center-plotly">
```{r}
#| label: by_region

modernization_by_region
```
</div>

## Democracy and Development Over Time

<br>

<div class="center-plotly">
```{r}
#| label: over_time2

modernization_over_time
```
</div>

# Discussion

## Three Theories

- Modernization; exogenous; and inclusive institutions (A&R)
- Sketch the causal arrow for each theory
- Choose a country (your favorite or from the list)

Possible cases (pick one): South Korea, Chile, Ghana, India, Singapore, Nigeria, Poland, Mexico, China

## Discussion Prompts

- Which theory fits your case best? Which fits worst?
- What evidence would change your mind?
- Is it "deviant" relative to any theory?
- How does your case fit to the "universe" of cases?
  - e.g. all of the countries for all of the years
